Retrofitting Effect Handlers onto OCaml

### 2020-11-19 retro-concurrency.tex
proof-reading edits

Figure~\ref{fig:fiber} shows the layout of a fiber in Multicore OCaml. At the
bottom of the stack, we have the |handler_info|, which contains the pointer to
parent fiber, and the closures for the value, exception and effect cases. The
closures are created by the translation described in
\S\ref{sec:static_semantics}; Multicore OCaml supports exception patterns in
addition to effect patterns in the same handler. This is followed by a context
block needed for DWARF and GC bookkeeping with callbacks. Then, there is a
top-level exception handler frame that forwards exceptions to the parent fiber.
When the exceptions are caught by this handler, the control switches to the
parent stack, and the exception handler closure |clos_hexn| is invoked. This is
followed by the |pc| of the code that returns values to the parent
fiber. This stack is laid out such that when the handled computation returns,
the control switches to the parent fiber and the value handler |clos_hval|
is invoked.

@@

When the effect |E| performed, we allocate a continuation object |ke| in the
heap that points to the current fiber |f|, set fiber |f|'s parent pointer to
|NULL|, and evaluate the continuation closure |clos_heffect| on the parent
fiber |e| with the effect |E| and the continuation |ke| as arguments. Since the
first handler does not handle effect |E|, the effect is \emph{reperformed}
(\S\ref{sec:static_semantics}) by appending the fiber |e| to the tail of
continuation |ke|, set fiber |e|'s parent pointer to |NULL|, and evaluate the
current continuation closure on the parent fiber |main| with |E| and |ke| as
arguments, which handles |E| (position |p2|). Thus, \emph{continuations are
captured without copying frames}. Since every handler closure is evaluated
until a matching one is found, the time taken to handle an effect is linear in
the number of handlers. We observed that the handler stack is shallow in
real programs.

@@

In this section, we evaluate the performance of Multicore OCaml effect handlers
against the performance requirements set in \S\ref{sec:req}. Multicore OCaml is
an extension of the 4.10.0 OCaml compiler with support for shared memory
parallelism and effect handlers. These results were obtained on 2-socket,
Intel\textregistered Xeon\textregistered Gold 5120 x86-64 server, with 28
physical cores (14 cores on each socket), and 2 hardware threads per core. Each
core runs at 2.20GHz and has 32 KB of L1 data cache, 32 KB of L1 instruction
cache and 1MB of L2 cache. The cores on a socket share a 19.25 MB L3 cache. The
server has 64GB of main memory and runs Ubuntu 18.04.

@@

In this section, we measure the impact of the addition of effect handlers on
code that does not use effect handlers. Our macro benchmark suite consists of
54 real OCaml workloads including verification tools (|Coq|, |Cubicle|,
|AltErgo|), parsers (|menhir|, |yojson|), storage engines (|irmin|), utilities
(|cpdf|, |decompress|), bioinformatics (|fasta|, |knucleotide|, |revcomp2|,
|regexredux2|), numerical analysis (|grammatrix|, |LU_decomposition|) and
simulations (|nbody|, |game_of_life|). In addition to Stock (|stock|) and
Multicore OCaml (|MC|), we also ran the benchmarks on Multicore OCaml with no
red zone (|MC+RedZone0|) in the fiber (all OCaml functions will have a stack
overflow check) and a red zone size of 32 words (|MC+RedZone32|). Recall that
the default red zone size in Multicore OCaml is 16 words (\S\ref{sec:hafibers}).

@@

We also present micro benchmarks results in Table~\ref{tab:micro_noeffect}.
Since micro benchmarks magnify micro-architectural optimisations, we also
report the number of instructions executed (obtained using |perf|) along with
time. |exnval| perform 100 million iterations of installing exception handlers
and returning with value. |exnraise| is similar, but raises an exception in
each iteration. |extcall| and |callback| perform 100 million external calls and
callbacks to identity functions. The other micro benchmarks are highly
recursive programs and were taken from ~\cite{Farvardin20}. For micro benchmarks,
we observed that padding tightloops with a few |nop| instructions, which
changes the loop alignment, makes the code up to 15\% faster. Hence, the
difference in running times under 15\% may not be statistically significant.

@@

Next, we aim to quantify the overhead of setting up and tearing down effect
handlers compared to a non-tail function call. To this end, we surround the
non-tail calls in the recursive micro benchmarks with an effect handler. These
programs do not |perform| effects. We also include a version of the same
benchmarks implemented using concurrency monad~\cite{Claessen99} (|monad|) as a
proxy for CPS versions. Recall that OCaml compiler does not use CPS in its IR.
In the |monad| version, we use a |fork| to invoke the non-tail call and use an
|MVar| to collect its result.

The results are presented in Table~\ref{tab:micro_noperform}. They show that
using effect handlers (concurrency monad) is 10.02$\times$ (67.09$\times$) more
expensive than the idiomatic implementation using non-tail calls. The concurrency
monad suffers due to the heap allocation of continuation frames (which need to
be GCed), whereas effect handlers benefit from stack allocation of the frames.
Our concurrency monad (and other monadic concurrency libraries such as
Lwt~\cite{lwt} and Async~\cite{async}) also have other downsides -- due to the
lack of a stack, exceptions, backtraces, and DWARF unwinding are no longer
useful.

@@

\textbf{\textit{Web server.}} As one of our primary goals of adding effect
handlers to OCaml is to support direct-style concurrency, we implement a
full-fledged HTTP/1.1 web server using effect handlers (|MC|). We use effect
handlers to implement a user-level threading library that spawns a thread per
request. We use \texttt{httpaf}~\cite{httpaf} for HTTP handling, and
\texttt{libev}~\cite{libev} for the eventloop. We compare our implementation
against an Lwt version (|lwt|) which also uses \texttt{httpaf} and
\texttt{libev}. Unlike using effect handlers, the Lwt version is written in monadic
style and does not have a notion of thread per request. For comparison, we
also include a Go 1.13 version (|go|) that uses the
\texttt{net/http}~\cite{nethttp} package. As both the OCaml versions are single
threaded, the Go benchmark is run with \texttt{GOMAXPROCS=1}.

@@

There are several strategies for implementing effect handlers. Eff~\cite{Eff},
Helium~\cite{Biernacki20}, Frank~\cite{Frank} and the Links server
backend~\cite{Hillerstrom20} use an interpreter similar to our operational
semantics to implement effect handlers. Effekt~\cite{Effekt}, Links JavaScript
backend~\cite{Hillerstrom20} and Koka~\cite{Leijen17} use type-directed
selective CPS translation. These language are equipped with an effect system,
which allows compiling pure code in direct-style and effectful code in CPS.
Leijen~\cite{Leijen14} implements effect handlers as a library in C using stack
copying. Since C allows pointers into the stack, care is taken to ensure that
when continuations are resumed, the constituent frames are restored to the same
memory addresses as at the time of capture. Kiselyov et al.~\cite{Kiselyov18}
use an implementation of multi-prompt delimited continuations as an OCaml
library~\cite{Kiselyov12} to embed Eff language in OCaml. Indeed, Forester et
al.~\cite{Forster19} showed that in an untyped setting, effect handlers,
monadic reflection and delimited control can macro-express each other.

>>>>

Figure~\ref{fig:fiber} shows the layout of a fiber in Multicore OCaml. At the
bottom of the stack, we have the |handler_info|, which contains the pointer to
the parent fiber, and the closures for the value, exception and effect cases.
The closures are created by the translation described in
\S\ref{sec:static_semantics}; Multicore OCaml supports exception patterns in
addition to effect patterns in the same handler. This is followed by a context
block needed for DWARF and GC bookkeeping with callbacks. Then, there is a
top-level exception handler frame that forwards exceptions to the parent fiber.
When the exceptions are caught by this handler, the control switches to the
parent stack, and the exception handler closure |clos_hexn| is invoked. This is
followed by the |pc| of the code that returns values to the parent
fiber. This stack is laid out such that when the handled computation returns,
the control switches to the parent fiber and the value handler |clos_hval|
is invoked.

@@

When the effect |E| is performed, we allocate a continuation object |ke| in the
heap that points to the current fiber |f|, set fiber |f|'s parent pointer to
|NULL|, and evaluate the continuation closure |clos_heffect| on the parent
fiber |e| with the effect |E| and the continuation |ke| as arguments. Since the
first handler does not handle effect |E|, the effect is \emph{reperformed}
(\S\ref{sec:static_semantics}) by appending the fiber |e| to the tail of
continuation |ke|, set fiber |e|'s parent pointer to |NULL|, and evaluate the
current continuation closure on the parent fiber |main| with |E| and |ke| as
arguments, which handles |E| (position |p2|). Thus, \emph{continuations are
captured without copying frames}. Since every handler closure is evaluated
until a matching one is found, the time taken to handle an effect is linear in
the number of handlers. We observed that the handler stack is shallow in
real programs.

@@

In this section, we evaluate the performance of Multicore OCaml effect handlers
against the performance requirements set in \S\ref{sec:req}. Multicore OCaml is
an extension of the 4.10.0 OCaml compiler with support for shared memory
parallelism and effect handlers. These results were obtained on a 2-socket,
Intel\textregistered Xeon\textregistered Gold 5120 x86-64 server, with 28
physical cores (14 cores on each socket), and 2 hardware threads per core. Each
core runs at 2.20GHz and has 32 KB of L1 data cache, 32 KB of L1 instruction
cache and 1MB of L2 cache. The cores on a socket share a 19.25 MB L3 cache. The
server has 64GB of main memory and runs Ubuntu 18.04.

@@

In this section, we measure the impact of the addition of effect handlers on
code that does not use effect handlers. Our macro benchmark suite consists of
54 real OCaml workloads including verification tools (|Coq|, |Cubicle|,
|AltErgo|), parsers (|menhir|, |yojson|), storage engines (|irmin|), utilities
(|cpdf|, |decompress|), bioinformatics (|fasta|, |knucleotide|, |revcomp2|,
|regexredux2|), numerical analysis (|grammatrix|, |LU_decomposition|) and
simulations (|nbody|, |game_of_life|). In addition to Stock (|stock|) and
Multicore OCaml (|MC|), we also ran the benchmarks on Multicore OCaml with no
red zone (|MC+RedZone0|) in the fibers (all OCaml functions will have a stack
overflow check) and a red zone size of 32 words (|MC+RedZone32|). Recall that
the default red zone size in Multicore OCaml is 16 words (\S\ref{sec:hafibers}).

@@

We also present micro benchmarks results in Table~\ref{tab:micro_noeffect}.
Since micro benchmarks magnify micro-architectural optimisations, we also
report the number of instructions executed (obtained using |perf|) along with
time. |exnval| performs 100 million iterations of installing exception handlers
and returning with a value. |exnraise| is similar, but raises an exception in
each iteration. |extcall| and |callback| perform 100 million external calls and
callbacks to identity functions. The other micro benchmarks are highly
recursive programs and were taken from ~\cite{Farvardin20}. For micro benchmarks,
we observed that padding tight loops with a few |nop| instructions, which
changes the loop alignment, can make the code up to 15\% faster. Hence, the
difference in running times under 15\% may not be statistically significant.

@@

Next, we aim to quantify the overhead of setting up and tearing down effect
handlers compared to a non-tail function call. To this end, we surround the
non-tail calls in the recursive micro benchmarks with an effect handler. These
programs do not |perform| effects. We also include a version of the same
benchmarks implemented using concurrency monad~\cite{Claessen99} (|monad|) as a
proxy for CPS versions. Recall that the OCaml compiler does not use CPS in its IR.
In the |monad| version, we use a |fork| to invoke the non-tail call and use an
|MVar| to collect its result.

The results are presented in Table~\ref{tab:micro_noperform}. They show that
using effect handlers (concurrency monad) is 10.02$\times$ (67.09$\times$) more
expensive than the idiomatic implementation using non-tail calls. The concurrency
monad suffers due to the heap allocation of continuation frames (which need to
be garbage collected), whereas effect handlers benefit from stack allocation of
the frames. Our concurrency monad (and other monadic concurrency libraries such
as Lwt~\cite{lwt} and Async~\cite{async}) also have other downsides -- due to
the lack of a stack, exceptions, backtraces, and DWARF unwinding are no longer
useful.

@@

\textbf{\textit{Web server.}} As one of our primary goals of adding effect
handlers to OCaml is to support direct-style concurrency, we implement a
full-fledged HTTP/1.1 web server using effect handlers (|MC|). We use effect
handlers to implement a user-level threading library that spawns a thread per
request. We use \texttt{httpaf}~\cite{httpaf} for HTTP handling, and
\texttt{libev}~\cite{libev} for the eventloop. We compare our implementation
against an Lwt version (|lwt|) which also uses \texttt{httpaf} and
\texttt{libev}. Unlike using effect handlers, the Lwt version is written in monadic
style and does not have the notion of a thread per request. For comparison, we
also include a Go 1.13 version (|go|) that uses the
\texttt{net/http}~\cite{nethttp} package. As both the OCaml versions are single
threaded, the Go benchmark is run with \texttt{GOMAXPROCS=1}.

@@

There are several strategies for implementing effect handlers. Eff~\cite{Eff},
Helium~\cite{Biernacki20}, Frank~\cite{Frank} and the Links server
backend~\cite{Hillerstrom20} use an interpreter similar to our operational
semantics to implement effect handlers. Effekt~\cite{Effekt}, Links JavaScript
backend~\cite{Hillerstrom20} and Koka~\cite{Leijen17} use type-directed
selective CPS translation. These language are equipped with an effect system,
which allows compiling pure code in direct-style and effectful code in CPS.
Leijen~\cite{Leijen14} implements effect handlers as a library in C using stack
copying. Since C allows pointers into the stack, care is taken to ensure that
when continuations are resumed, the constituent frames are restored to the same
memory addresses as at the time of capture. Kiselyov et al.~\cite{Kiselyov18}
use an implementation of multi-prompt delimited continuations as an OCaml
library~\cite{Kiselyov12} to embed the Eff language in OCaml. Indeed, Forester et
al.~\cite{Forster19} showed that in an untyped setting, effect handlers,
monadic reflection and delimited control can macro-express each other.
